---
title: 'A2: Explore Gapminder and Use dplyr'
output: github_document
---

```{r, echo=T, include=F}
library(gapminder)
library(dplyr)
library(tidyverse)
library(datasets)
```

# Exercise 1: Basic `dplyr`
##1.1
```{r}
cuj <- filter(gapminder, year %in% seq(1970,1979,1), 
        country == "China" | country ==  "United States" | country == "Japan")
cuj
```

##1.2
```{r}
cuj %>% 
  select(country, gdpPercap)
```

##1.3
```{r}
inc.in.lifeExp <- c(diff(gapminder$lifeExp), NA) #create a new variable of difference
gapminder %>%  mutate(inc.in.lifeExp) %>%  #add variable as a column into gapminder
  filter(inc.in.lifeExp < 0) #discard entries with a positive difference 
```

##1.4
```{r}
maxgdata <- gapminder %>% 
  group_by(country) %>% 
  filter(gdpPercap == max(gdpPercap))
maxgdata
```
## 1.5
```{r}
filter(gapminder, country %in% "Canada") %>% 
  select(country, lifeExp, gdpPercap) %>% 
  ggplot(aes(x = lifeExp, y = log(gdpPercap))) + geom_point() +
  labs(title = "Life expectancy vs. log(GDP per capita) in Canada") +
  theme(plot.title = element_text(hjust = 0.5)) #centering the title
```

# Exercise 2
The categorical variable and quantative variable of choice are continent and life expectancy, respectively. 

**Q1: What are possible values (or range, whichever is appropriate) of each variable?**

**Solution:**

The possible values/groups of continent are Africa, Americas, Asia, Europe, and Oceania.

The range of life expectancy across all continents varies from 23.6 years old to 82.60 years old. 
```{r}
gapminder %>% group_keys(continent) #access possible values in continent
range(gapminder$lifeExp)
```

**Q2: What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.**

**Solution:**

Africa is the continent with the most information recorded in the gapminder dataset; specifically, it has 624 entries.

The most common value for life expectancy is 69.39 years old across all continents.

The spread of life expectancy data can be described by its IQR, standard deviation, and quantiles. For instance, its IQR is 22.6475 years old. Its standard deviation is 12.91711 years old. Lastly, the majority of the data lie between 48.198 and 70.8455 years old.

The life expectancy of Europeans and Africans contains many outliers, with that of the Europeans being the most. All continents except Africa and Oceania demonstrate left-skewness meaning that most people have a relatively high life expectancy. Oceanians' life expectancy is distributed approximately symmetrically without any outlier present. However, Africans' life expectancy is slightly skewed to the right suggesting that most people do not have a very high life expectancy.

```{r}
gapminder %>% count(continent)
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
getmode(gapminder$lifeExp)
sumtable <- as.table(summary(gapminder$lifeExp))
sumtable
gapminder$lifeExp %>% sd()
gapminder$lifeExp %>% quantile()
gapminder$lifeExp %>% IQR()
gapminder %>% ggplot(aes(x = continent, y = lifeExp)) + geom_boxplot() 
```

# Exercise 3
Below is a scatterplot depicting the relationship between petal length and petal width of different species of iris flowers from the `iris` dataset. Trend line with confidence region is also included for each species in order to generate a more clear trend.

We can see that the length and the width of iris petals are positively associated across all species. Furthermore, **virginica** appears to have the largest petals while **setosa** has the smallest overall.

```{r}
iris %>%
  ggplot(aes(x=Petal.Length, y=Petal.Width, color = Species)) +
  geom_point() + theme_minimal() + geom_smooth()
```

A boxplot of species versus sepal length can be examined to visualize the length distribution across different iris flowers.

```{r}
iris %>% 
  ggplot(aes(x=Species, y=Sepal.Length,color=Species)) +
  geom_boxplot() +theme_minimal() + theme(legend.position="right")
```

# Recycling

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

Now, let's filter the observations individually to check if the analyst included all relevant observations.

```{r}
filter(gapminder, country == "Rwanda")
filter(gapminder, country == "Afghanistan")
```

The analyst only managed to produce 12 observations in total. However, it is clear that Rwanda and Afghanistan each has 12 observations. Therefore, the analyst did *not* suceed.

Below code would fix the analyst's error:
```{r}
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

