---
title: "Assignment 4: Tidy data and joins"
output:
  html_document:
    toc: true
---
```{r, warning = FALSE, message = FALSE, include=F}
library(tidyverse)
library(gapminder)
library(ggplot2)
library(readr)
```

#Exercise 1: Univariate Data Reshaping
## 1.1
Make a tibble with one row per year, and columns for life expectancy for two or more countries.
```{r}
gapwide <- gapminder %>% 
  filter(country == "Afghanistan" | country == "Algeria") %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp)
DT::datatable(gapwide)
```

##1.2
Take advantage of this new data shape to make a scatterplot of life expectancy of one country against that of another.

```{r}
gapwide %>% 
  ggplot(aes(x=Afghanistan, y=Algeria)) +
  geom_point(size=2, shape=23) +
  labs(title = "Life Expectancy of Afghanistanis vs that of Algerians in 1952-2007",
       x = "Afghanistanis' Age (years)", y = "Algerians' Age (years)" ) +
  theme(plot.title = element_text(hjust = 0.5))
```

##1.3
Re-lengthen the dataset
```{r}
gapwide %>% 
  pivot_longer(cols = -year,
               names_to = "country",
               values_to = "life.exp") %>% 
  arrange(country) %>% 
  DT::datatable()
```
 
#Exercise 2: Multivariate Data Reshaping

##2.1
Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.
```{r}
multigapwide <- gapminder %>% 
  filter(country == "Afghanistan" | country == "Algeria") %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(gdpPercap, lifeExp))
DT::datatable(multigapwide)
```
 

##2.2
Re-lengthen the data.
```{r}
multigapwide %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_" ) %>% 
  arrange(country) %>% 
  DT::datatable()
```

#Exercise 3: Table Joins

```{r, warning = FALSE, message = FALSE, echo = FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```
 
##3.1
For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.

```{r}
colnames(email)[colnames(email) == "guest"] <- "name" #make name column same across datasets
guest %>% 
  left_join(email, by = "name") %>% 
  DT::datatable()
```

##3.2
Who do we have emails for, yet are not on the guestlist?

```{r}
emailnew <- separate_rows(email, name, sep = ", ", convert = T) #separate guests' names
emailnew %>% 
  anti_join(guest, by = "name") %>% 
  DT::datatable()
```
Turner, Albert, and Vivian all have emails but not on the guestlist. 

##3.3
Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

```{r}
guest %>% 
  full_join(emailnew, by = "name") %>% 
  DT::datatable()
```




